cardinalR: Generating interesting high-dimensional data structures

### 2025-05-15 paper-cardinalR.Rmd
Updated text in
Removed data sets section

## Data sets

The `cardinalR` package comes with several data sets that load with the package. These are described in Table `r knitr::asis_output(ifelse(knitr::is_html_output(), '\\@ref(tab:datasets-tb-html)', '\\@ref(tab:datasets-tb-pdf)'))`.

```{r}
datasets_tb <- tibble(dt = c("mobiusgau", 
                             "mobiusgau_tsne1", 
                             "mobiusgau_tsne2",
                             "mobiusgau_tsne3",
                             "mobiusgau_umap1",
                             "mobiusgau_umap2",
                             "mobiusgau_umap3"), 
                      text = c("Simulated data with a Mobius and a Gaussian in 4-D space.",
                               "The tSNE embedding with perplexity $15$ for mobiusgau.", 
                               "The tSNE embedding with perplexity $30$ for mobiusgau.",
                               "The tSNE embedding with perplexity $5$ for mobiusgau.",
                               "The UMAP embedding with number of neighbors $15$ and minimum distance $0.1$ for mobiusgau.",
                               "The UMAP embedding with number of neighbors $30$ and minimum distance $0.08$ for mobiusgau.", 
                               "The UMAP embedding with number of neighbors $5$ and minimum distance $0.9$ for mobiusgau."))
```

```{r datasets-tb-html, eval=knitr::is_html_output()}
datasets_tb |> 
  kable(caption = "cardinalR data sets", col.names = c("data", "explanation")) 
```

```{r datasets-tb-pdf, eval=knitr::is_latex_output()}
datasets_tb |> 
  kable(caption = "cardinalR data sets", format="latex", col.names = c("data", "explanation"), booktabs = T)  |>
  column_spec(1, width = "4cm") |>
  column_spec(2, width = "8cm")
```

@@

Each branch is a segment of a line with added jitter to simulate measurement noise. The branches differ in direction and location. Branches 1 and 2 are initialized with fixed slopes and intercepts. The Branch 1 is generated from $X_1 \sim U(-2, 8)\$, \$X_2 = 0.5 \cdot X_1 + \epsilon$. The Branch 2 is generated from $X_1 \sim U(-6, 2)\$, \$X_2 = -0.5 \cdot X_1 + \epsilon$. $\epsilon \sim U(0, 0.5)$.

Branches 3 to $k$** are added iteratively. Each starts at a location outside predefined exclusion zones to avoid overlap with the initial two branches.

@@

Branches $3$ to $k$** are added iteratively. Each new branch begins at a randomly chosen point within a restricted horizontal band: $X_{1} \in [-0.15, 0.15]$ (to ensure connectivity with earlier branches), spans a unit-length interval on $X_{1}$:$X_1 \in [x_{\text{start}}, x_{\text{start}} + 1]$, and has the structure:

@@

* If $k \leq \binom{p}{2}$: $s_i = 1 $
* If $k > \binom{p}{2}$: $s_i \sim \text{Uniform sample from } {1, 1.5, \dots, 8}$

@@

The shape is generated using a third-degree raw polynomial basis expansion. Let $X_1 \sim U(\text{range}[1], \text{range}[2])$ be the base input. A cubic transformation with vertical jitter is used to define $X_2 = X_1 + X_1^2 - X_1^3 + \varepsilon_2\$, where \$\varepsilon_2 \sim U(0, 0.5)$.

>>>>

@@

Each branch is a segment of a line with added jitter to simulate measurement noise. The branches differ in direction and location. Branches 1 and 2 are initialized with fixed slopes and intercepts. The Branch 1 is generated from $X_1 \sim U(-2, 8)$, $X_2 = 0.5 \cdot X_1 + \epsilon$. The Branch 2 is generated from $X_1 \sim U(-6, 2)$, $X_2 = -0.5 \cdot X_1 + \epsilon$. $\epsilon \sim U(0, 0.5)$.

Branches 3 to $k$ are added iteratively. Each starts at a location outside predefined exclusion zones to avoid overlap with the initial two branches.

@@

Branches $3$ to $k$ are added iteratively. Each new branch begins at a randomly chosen point within a restricted horizontal band: $X_{1} \in [-0.15, 0.15]$ (to ensure connectivity with earlier branches), spans a unit-length interval on $X_{1}$:$X_1 \in [x_{\text{start}}, x_{\text{start}} + 1]$, and has the structure:

@@

* If $k \leq \binom{p}{2}$: $s_i = 1$
* If $k > \binom{p}{2}$: $s_i \sim \text{Uniform sample from } {1, 1.5, \dots, 8}$

@@

The shape is generated using a third-degree raw polynomial basis expansion. Let $X_1 \sim U(\text{range}[1], \text{range}[2])$ be the base input. A cubic transformation with vertical jitter is used to define $X_2 = X_1 + X_1^2 - X_1^3 + \varepsilon_2$, where $\varepsilon_2 \sim U(0, 0.5)$.
